\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[a4paper, top=25mm, bottom=25mm, left=25mm, right=25mm]{geometry}
\usepackage{pgfplots}
\usepackage{mathtools}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{fillbetween}
\usepackage{comment}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{newunicodechar}
\newunicodechar{₺}{\ensuremath{\mathrm{TL}}}

\begin{document}

\large

\begin{center}
{\huge MAT123 07.11.2025}
\end{center}

\hfill

{\Large \textbf{QUESTIONS}}

\hfill

\noindent \textbf{Q1:} For $0 < t \le 1$, let $A(t)$ denote the area of the triangle bounded by the $x$-axis, the $y$-axis and the tangent line to the curve $y = \ln x$ at $(t, \ln t)$. Find the maximum value of $A(t)$.

\hfill

\noindent \textbf{Q2:} A manufacturer will construct an open rectangular box with a square base and a volume of $6\,\text{m}^3$. If material for the bottom costs 60 Turkish Liras/m$^2$ and material for the sides costs 40 Turkish Liras/m$^2$, what dimensions will result in the least expensive box? What is the minimum cost?

\hfill

\noindent \textbf{Q3:} Find $y'(1)$ if $x^y + y^x = 3$.

\hfill

\noindent \textbf{Q4:} Show that the equation $x^3+x^2+2x+5=0$ has a root, but no more than one.

\hfill

\noindent \textbf{Q5:} Let $C$ be the graph of the equation $3y^2x-x^3-y^3-y=0$. Find all the points on $C$ through which the tangent lines to $C$ are horizontal.

\hfill

\noindent \textbf{Q6:} How fast is the surface area of a cube changing when the volume of the cube is $64$ cm$^2$ and is increasing at $2$ cm$^3$/sec?

\hfill

\noindent \textbf{Q7:} Evaluate $\displaystyle\lim_{x\to\infty}\left(\sqrt{x+\sqrt x}-\sqrt x\right)$.

\hfill

\noindent \textbf{Q8:} Let $f(x)=\dfrac{x^2-2}{x^2-1}$.

\hfill

\textbf{a}. Find the entire domain.

\textbf{b}. Find the $x$- and $y$- intercepts.

\textbf{c}. Find the asymptotes.

\textbf{d}. Determine $f'(x)$ and $f''(x)$.

\textbf{e}. Make a table.

\textbf{f}. Sketch the graph.

\hfill

\noindent \textbf{Q9:} Find $\dfrac{dy}{dx}$ where $y=(\ln x)^{\ln x}$
 
\hfill

\noindent \textbf{Q10:} Evaluate $\displaystyle\lim_{x\to1^-}\frac{\ln\sin\pi x}{\csc\pi x}$.

\hfill

\hfill

{\Large \textbf{ANSWERS}}

\hfill

\noindent \textbf{Q1:} $\dfrac 2e$ \hspace{30pt}\textbf{Q2:} Dimensions: $2\text{m} \times \text{2m} \times \text{1.5m}$, Cost: 720 TL
 \hspace{30pt} \textbf{Q3:} $y=-2\ln2-2$.

\hfill

\noindent \textbf{Q4:} Proof by means of IVT and MVT. \hspace{15pt}\textbf{Q5:} $\{(-1,-1),(0,0),(1,1)\}$ \hspace{15pt}\textbf{Q6:} 2 cm$^2/$sec

\hfill

\noindent \textbf{Q7:} $\dfrac12$

\newpage

\noindent \textbf{Q8:} \textbf{a.} $\mathcal{D}_f=\mathbb R\setminus\{-1,1\}$ \hspace{15pt} \textbf{b.} $\{(0,2),(-\sqrt2,0),(\sqrt2,0)\}$

\hfill

\noindent \textbf{c.} Horizontal: $y=1$, vertical: $x=\pm1$\hspace{15pt}\textbf{d.} $f'(x)=\dfrac{2x}{(x^2-1)^2},\quad f''(x)=\dfrac{-6x^2-2}{(x^2-1)^3}$

\hfill

\noindent \textbf{e. \& f.} on the paper below.

\hfill

\noindent \textbf{Q9:} $y'=(\ln x)^{\ln x}\left[\dfrac1x\ln(\ln x)+\dfrac1x\right]$\hspace{15pt}\textbf{Q10:} 0

\hfill

\end{document}